Normalization by Evaluation for Call-by-Push-Value

### 2019-04-01 fscd19/response-fscd19.txt

A big thanks to the reviewers who spent considerable time to read and
understand our paper and to provide valuable feedback and pointers to
relevant literature.  We will take all comments into account when
preparing the revision.

In the following, we clarify the main questions of the reviewers:

@@

Without computational soundness, there was no need to retain the
equational theory of IPL, thus, we relegated it to the appendix.
Indeed, our normal forms can be understood purely logically, based on
the distinction of introductions from eliminations and negative
(choiceless) types from positive (choiceful) types.

- Missing theorems (reviewers 1,3)

- Incomplete formalization (reviewers 1,2)

Reviewers 2 and 3 note that the Agda formalization is incomplete and
unpolished.  The Agda formalization helped us to get the constructions
right.  However, we like to think of this paper as an implementation
of NbE in *Type Theory*.  Systems like Agda disappear (where is LEGO
now?), but type theory is likely to stay with us for a while.  Thus,
we do not consider the Agda code as a contribution of this paper.
Instead, we have striven for a rigorous, self-contained mathematical
presentation.

@@

In the context of focused lambda-calculi, which we study to reduce the
degrees of liberty in normal forms (in other words, removing
don't-care non-determinism), CBPV arises naturally as calculus with
focusing on positive introductions but not on positive eliminations.
By "coincidence" (or rather not by coincidence, since there are no
coincidences in mathematics) it is the same calculus as the one found
in the search for a lambda-calculus with effects.

We are indeed only interested in CBPV as logical system.  Normalizing
effects is beyond the scope of this paper.  In any case, since effects
do not follow the introduction/elimination scheme, additional
considerations would be necessary for normalizing effects and would
like likely depend on the individual effects considered.

- Future work: relationship to CBN and CBV translations (reviewers 1,2)

We are sorry we could not satisfy the curiosity of our reviewers at
this point.  The idea to investigate the relationship to the CBN/CBV
translations came late in the process of writing, and there was
neither time nor space to address it.

We hope that we have addressed the major questions of the reviewers.
Replies to many minor questions can be found in the longer version at

  https://github.com/andreasabel/ipl/blob/master/fscd19/response-fscd19.txt

> ----------------------- REVIEW 1 ---------------------
> PAPER: 50
> TITLE: Normalization by Evaluation for Call-by-Push-Value and Polarized Lambda-Calculus
> AUTHORS: Andreas Abel and Christian Sattler
>
> Overall evaluation: 0 (borderline paper)
>
> ----------- Overall evaluation -----------
>
> This paper present a definition of normalization by evaluation for
> call-by-push-value.  The authors first present NbE for a simply type
> lambda calculus with sums. For that, they need to introduce a
> (coverage) monad to be able to do backtracking or to build a case tree
> of possible normalization. A version of this monad is then used to
> interpret computations in CBPV and gives rise to an NbE algorithm for
> CBPV. Finally, the authors discuss modification for focused polarized
> logic, which is closed to CBPV.
>
> General Comment :
>
> This is an interesting work on how to extend NbE technique that are
> compatible with the introduction of effect in the calculus. The paper
> seems mathematically correct and the idea of using such a monad to
> extend NbE to potentially effectful terms is very natural.
>
> However, I have severe reserves about the paper. Therefore I cannot be
> positive about this paper.
>
> - The paper is missing theorem in two senses. First, it is missing a
>   clear statement of what is claimed to be achieved, namely “logical
>   consistency”. The expected theorem are not there.  I was expected at
>   least computational soundness. Also I found quite disappointing that
>   the the combination with the CBN and CBV translation to CBPV has not
>   been investigated, without even explaining why. No time, too
>   difficult, not sure is corresponds to usual NbE ?

@@

> - The formalization in Agda i just a link to a GitHub repo that is
>   very hard to dive into. There is no README to know what has been
>   done and where is has been done.  We don’t know what has been
>   formalized in Agda for section 4. It seems that nothing has been
>   formalized for section 3, why ?  The repo looks more like the link
>   to an ongoing project than a link to some documented and finish
>   material aimed at being understood by a reader.

The Agda formalization helped us getting the constructions correct,
but the paper is intended to be self-contained without the Agda
formalization.

> Minor remarks:
>
> The paper is longer than the 15 pages including biblio of the CFP, but
> I didn’t take this into account in my review.
>
> However, putting the equational theory in Appendix rather feels like
> the authors needed to save more space and decided quite arbitrarily to
> cut this.
>
> In the definition of ↑A⇒B ,  where does ∆ comes from ?

When quantifying over (τ : Γ ⊆ Δ) we implicitly quantify over Δ,
similarly as when writing ∀ Δ ⊇ Γ. P, which is a common shorthand for
∀ Δ. Γ ⊆ Δ ⇒ P.

@@

> 519: What's hard about the agreement proof? Or do you just leave it
> out because you think it's out of scope for the paper?

We are already over the page limit...

> Minor comments
>
> 119: I would maybe write "x : (A ∈ Γ)" to make clear that "x : A ∈ Γ"
> is not a definition here, but you really mean an inhabitant x of a
> type. In named presentations it might be a short hand for "Γ x = A"

See 105-109 for explicit explanations of ":" and "∈".
See also "suc zero : A ∈ Γ.A.B" in line 118.
WON'T FIX.

> 122: Is Cxt a typo? If not, a sentence why Ctx is not used might make
> sense. Is it to avoid confusion with the usual presentation of Ctx
> with arrows inverse to Cxt?

@@

The jargon is indeed not necessary, but it connects to the categorical
notions.  The paper is intended as a bridge between conventional and
categorical meta-theory of lambda-calculus, and we want to be
accessible for both camps.

> No theorems.  The paper presents no formal definitions or theorems
> but simply develops a narrative.  I guess the existence of certain
> dependently typed functions takes the pace of any formal theorems,
> but for me it would still have been helpful to see standard
> theorems regarding the properties of various constructions.  It
> might also help to put some of the definitions into figures so
> they can be referenced in the text (even if the authors want to
> avoid explicit Definition 1, Definition 2, ...).

"No formal definitions" is probably not what was meant here, since, in
fact, *all* definitions in the paper are fully formal (with a few
definitions omitted).
Indeed, as you say, dependently typed total functions second as proofs
of the theorems their types express.  Since we work in Martin-Löf Type
Theory, all are constructions are proof-relevant, thus, there are no
theorems in the original sense.

All constructions we give are uniquely determined by their name, thus,
there is no secondary reference structure such as Definition nnn.  To
help the reader with locating the definition, we have boxed important
definitions which do not have their own display.

### 2019-05-05 types19/types19.tex
[ TYPES 2019 ] formatted down to 2 pages (condensed bib/ackn)

\usepackage{natbib}
\usepackage{hyperref}
% \usepackage{agda} % colors

@@

@@

\title{Normalization by Evaluation for Call-by-Push-Value}
\titlerunning{Normalization by Evaluation for Call-by-Push-Value}
\authorrunning{A. Abel}
\author{Andreas Abel}
\institute{
  Department of Computer Science and Engineering,
  Gothenburg University
}

@@

NbE for sum types requires a refinement of the model, since reflection of
a variable of type $A + B$ as a value in $\den{A + B}$ requires case
distinction in the model.  One such refinement are \emph{sheaves}
\citep{altenkirchDybjerHofmannScott:lics01}; another is the use of a
monad $\C$
\citep{filinski:semaccounttdpe,filinski:tlca01,barral:PhD}
in the category of presheaves
for the interpretation of sum types:
$\den{A + B} = \C(\den A + \den B)$.
The smallest such \emph{``cover'' monad} $\C$ are binary trees where leaves are
the monadic unit aka $\treturn$, and the nodes case distinctions
over neutrals $\tNe\,(A_1 + A_2)$ of sum type.  When leaves are normal
forms, the whole tree represents a normal form, thus,
$\trunNf : \C(\tNf\,A) \to \tNf\,A$ is trivial.
This \emph{running of the monad}
on normal forms represents the algorithmic part of the sheaf condition
on $\tNf\,A$ and extends as $\trun : \C \den A \to \den A$ to all
semantic types.

@@

CBPV was designed to study lambda-calculus with effects.  It separates
types into \emph{value} types $P$ and \emph{computation} types $N$, which we, in
analogy to polarized lambda-calculus
\citep{zeilberger:PhD,brockNannestadSchuermann:lpar10,espiritoSanto:entcs17}
refer to as \emph{positive} and \emph{negative} types.
Variables stand for values, thus, have positive types.
The monad that models the effects is placed at the transition from values
to computations $\Comp P$, and computations can be embedded into
values by \emph{thunking} ($\Thunk N$).
\[
\begin{array}{lllrl@{\qquad}l}
  \PTy & \ni & P & ::= & \patom
    % \mid \tyone \mid P_1 \times P_2
    % \mid \tyzero
    \mid P_1 + P_2
    \mid \Thunk N
  & \mbox{positive type / value type}
\\
  \NTy & \ni & N & ::= & % \natom
    % \mid \top \mid N_1 \ntimes N_2
    % \mid
    P \To N
    \mid \Comp P
  & \mbox{negative type / computation type}
\\
\end{array}
\]
We restrict to a fragment of \emph{pure} CBPV
with a single positive connective, sum
types $P_1 + P_2$, and a single negative connective, call-by-value
function types $P \To N$.  While we have no proper effects, the
evaluation of open terms requires the effect of case distinction over
neutrals, modeled by a cover monad $\C$.  In the following, we give
inductive definitions of the presheaves of
normal ($\tNf$) and neutral normal forms
($\tNe$) of our fragment of CBPV and a concrete, strong cover monad $\Cov$.
\begin{small}
\begin{gather*}
  \nru{\tvar}{\Var \patom \Gamma}{\PValNf {\patom} \Gamma}
\quad
  \nru{\tthunk}{\Nf N \Gamma}{\PValNfP{\Thunk N}\Gamma}
\quad
  \nru{\tinj_i}{\PValNf{P_i}\Gamma}{\PValNfP{P_1 + P_2}\Gamma}
%\\[2ex]
\quad
  \nru{\tret}
      {\CoverP {\tVNf\,P}  \Gamma}
      {\NfP {\Comp P} \Gamma}
\quad
  \nru{\tabs}
      {\Nf N {(\cext \Gamma P)}}
      {\NfP {P \To N} \Gamma}
\end{gather*}
\end{small}
\begin{small}
\begin{gather*}
  \nru{\tforce}{\VarP {\Thunk N} \Gamma}{\Ne N \Gamma}
\qquad
  \nru{\tapp}
      {\NeP {P \To N} \Gamma \qquad \PValNf P \Gamma}
      {\Ne N \Gamma}
\qquad
  \nru{\tbind}
      {\NeP {\Comp P} \Gamma \qquad \Cover \J (\cext \Gamma P)}
      {\Cover \J \Gamma}
\\[2ex]
  \nru{\treturn}{\J\,\Gamma}{\Cover \J \Gamma}
\qquad
  \nru{\tcase}
      {\VarP {P_1 + P_2} \Gamma
       \qquad \Cover \J (\cext \Gamma {P_1})
       \qquad \Cover \J (\cext \Gamma {P_2})
      }
      {\Cover \J \Gamma}
\end{gather*}
\end{small}
\noindent
($\J$ stands for an arbitrary presheaf in $\Cov\,\J$.)
Normal forms start from a variable of base type and continue with
introductions,
except that the services of the monad can be used at the
transition $\tret$ from positive to negative types ($\Comp P$).
Neutrals are eliminations of variables of type $\Thunk N$ into a
positive type $\Comp P$, and can then be bound
to a variable of type $P$ to be used in a computation (see $\tbind$).
Variables of
sum type $P_1 + P_2$ can be utilized in computations through a
$\tcase$ split.

@@

\paragraph*{Acknowledgments.}
%Thanks to the anonymous referees, who helped improving the quality of
%this abstract through their feedback.
This work was supported by
Vetenskapsr\aa{}det under Grant
  No.~621-2014-4864 \emph{Termination Certificates
  for Dependently-Typed Programs and Proofs via Refinement Types}
and by the EU Cost Action CA15123 \emph{Types for programming and verification}.

@@

\bibliographystyle{abbrvnat}
\bibliography{auto-types19}

>>>>

\usepackage{natbib}
\usepackage{hyperref}
\urlstyle{sf}
% \usepackage{agda} % colors

@@

% less space between entries in bibliography
\let\OLDthebibliography\thebibliography
\renewcommand\thebibliography[1]{
  \OLDthebibliography{#1}
  \setlength{\parskip}{0pt}
  \setlength{\itemsep}{0.2ex plus 0.3ex}
}

@@

\title{Normalization by Evaluation for Call-by-Push-Value}
\titlerunning{Normalization by Evaluation for Call-by-Push-Value}
\authorrunning{A. Abel}
\author{Andreas Abel\thanks{Supported by VR grant 621-2014-4864 and EU
  Cost Action CA15123.}}
\institute{
  Department of Computer Science and Engineering,
  Gothenburg University
}

@@

NbE for sum types requires a refinement of the model, since reflection of
a variable of type $A + B$ as a value in $\den{A + B}$ requires case
distinction in the model.  One such refinement are \emph{sheaves}
\citep{altenkirchDybjerHofmannScott:lics01}; another is the use of a
monad $\C$
\citep{%filinski:semaccounttdpe,
filinski:tlca01,barral:PhD}
in the category of presheaves
for the interpretation of sum types:
$\den{A + B} = \C(\den A + \den B)$.
The smallest such \emph{``cover'' monad} $\C$ are binary trees where leaves are
the monadic unit aka $\treturn$, and the nodes case distinctions
over neutrals $\tNe\,(A_1 + A_2)$ of sum type.  When leaves are normal
forms, the whole tree represents a normal form, thus,
$\trunNf : \C(\tNf\,A) \to \tNf\,A$ is trivial.
This \emph{running of the monad}
on normal forms represents the algorithmic part of the sheaf condition
on $\tNf\,A$ and extends as $\trun : \C \den A \to \den A$ to all
semantic types.

@@

CBPV was designed to study lambda-calculus with effects.  It separates
types into \emph{value} types $P$ and \emph{computation} types $N$, which we, in
analogy to polarized lambda-calculus
\citep{zeilberger:PhD} %,brockNannestadSchuermann:lpar10,espiritoSanto:entcs17}
refer to as \emph{positive} and \emph{negative} types.
Variables stand for values, thus, have positive types.
The monad that models the effects is placed at the transition from values
to computations $\Comp P$, and computations can be embedded into
values by \emph{thunking} ($\Thunk N$).
\[
\begin{array}{lllrl@{\qquad}l}
  \PTy & \ni & P & ::= & \patom
    % \mid \tyone \mid P_1 \times P_2
    % \mid \tyzero
    \mid P_1 + P_2
    \mid \Thunk N
  & \mbox{positive type / value type}
\\
  \NTy & \ni & N & ::= & % \natom
    % \mid \top \mid N_1 \ntimes N_2
    % \mid
    P \To N
    \mid \Comp P
  & \mbox{negative type / computation type}
\\
\end{array}
\]
We restrict to a fragment of \emph{pure} CBPV
with a single positive connective, sum
types $P_1 + P_2$, and a single negative connective, call-by-value
function types $P \To N$.  While we have no proper effects, the
evaluation of open terms requires the effect of case distinction over
neutrals, modeled by a cover monad $\C$.  In the following, we give
inductive definitions of the presheaves of
normal ($\tNf$) and neutral normal forms
($\tNe$) of our fragment of CBPV and a concrete, strong cover monad $\Cov$.
\begin{small}
\begin{gather*}
  \nru{\tvar}{\Var \patom \Gamma}{\PValNf {\patom} \Gamma}
\quad
  \nru{\tthunk}{\Nf N \Gamma}{\PValNfP{\Thunk N}\Gamma}
\quad
  \nru{\tinj_i}{\PValNf{P_i}\Gamma}{\PValNfP{P_1 + P_2}\Gamma}
%\\[2ex]
\quad
  \nru{\tret}
      {\CoverP {\tVNf\,P}  \Gamma}
      {\NfP {\Comp P} \Gamma}
\quad
  \nru{\tabs}
      {\Nf N {(\cext \Gamma P)}}
      {\NfP {P \To N} \Gamma}
\end{gather*}
\end{small}
\begin{small}
\begin{gather*}
  \nru{\tforce}{\VarP {\Thunk N} \Gamma}{\Ne N \Gamma}
\qquad
  \nru{\tapp}
      {\NeP {P \To N} \Gamma \qquad \PValNf P \Gamma}
      {\Ne N \Gamma}
\qquad
  \nru{\tbind}
      {\NeP {\Comp P} \Gamma \qquad \Cover \J (\cext \Gamma P)}
      {\Cover \J \Gamma}
\\[2ex]
  \nru{\treturn}{\J\,\Gamma}{\Cover \J \Gamma}
\qquad
  \nru{\tcase}
      {\VarP {P_1 + P_2} \Gamma
       \qquad \Cover \J (\cext \Gamma {P_1})
       \qquad \Cover \J (\cext \Gamma {P_2})
      }
      {\Cover \J \Gamma}
\end{gather*}
\end{small}
\noindent
($\J$ stands for an arbitrary presheaf in $\Cov\,\J$.)
Normal forms start from a variable of base type and continue with
introductions,
except that the services of the monad can be used at the
transition $\tret$ from positive to negative types ($\Comp P$).
Neutrals are eliminations of variables of type $\Thunk N$ into a
positive type $\Comp P$, and can then be bound
to a variable of type $P$ to be used in a computation (see $\tbind$).
Variables of
sum type $P_1 + P_2$ can be utilized in computations through a
$\tcase$ split.

@@

% \paragraph*{Acknowledgments.}
% %Thanks to the anonymous referees, who helped improving the quality of
% %this abstract through their feedback.
% This work was supported by
% Vetenskapsr\aa{}det under Grant
%   No.~621-2014-4864 \emph{Termination Certificates
%   for Dependently-Typed Programs and Proofs via Refinement Types}
% and by the EU Cost Action CA15123 \emph{Types for programming and verification}.

@@

\bibliographystyle{abbrvnat}
%\bibliographystyle{abbrvurl}
%\bibliographystyle{plainurl} % no author-year cit.
\bibliography{auto-types19}
